---
title: "IODS final project"
author: "Otso Perakyla"
output:
  html_document:
    theme: journal
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

In this final assignment, I will be analysing data on carbon dioxide concentration from four measurement sites around the world. The data can be found from National Ocean and Atmosphere Administration (NOAA) [here](https://www.esrl.noaa.gov/gmd/dv/data/index.php?parameter_name=Carbon%2BDioxide&type=Insitu). The measurement sites I will analyze are Mauna Loa in Hawaii (MLO), Barrow in Alaska (BRW), Tutuila in American Samoa (SMO), and South Pole, Antarctica (SPO). Carbon dioxide is the main greenhouse gas responsible for anthropogenically enhanced greenhouse effect, and thus climate change. Therefore it is important to understand its behaviour in the atmosphere. I will try to analyze the rate of its increase, seasonal differences and the differences between the measurement stations. 

```{r load_data}


# clear all variables possibly left over from previous sessions
rm(list = ls())

# load libraries

library(GGally)
library(ggplot2)
library(dplyr)

# load the data
CO2_all <- read.table("./data/CO2_4_sites.txt",header = T,  dec = ".")

```

I have already wrangled the data into an easily analyzable form: the wrangling script can be found [here](https://github.com/otsopera/IODS-final/blob/master/data/create_CO2.R). The wrangled data contains `r nrow(CO2_all) ` monthly carbon dioxide concentration values from the stations. Some of the observations are missing. In total, the data contains `r ncol(CO2_all) ` variables, which are the following:

```{r vars}

variable.names(CO2_all)


```

Here the variable suffixes (_MLO, _BRW, _SMO, _SPO) tell the station where the measurement comes from. Tha variable value is the monthly mean carbon dioxide concentration, in parts per million (ppm). value_unc is the uncertainty estimate for the measurement, also in ppm. nvalue tells how many individual measurements have gone into the value: for one monthly value, multiple individual measurements have been averaged. Lastly, the date_time tells the month that the measurement corresponds to. 




## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}



qplot(date_time,value_MLO,data = CO2_all) 
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
